---
title: Ordinal Data Practical for IBG Boulder Workshop
author:
  - Mike Neale
  - Liz Prom-Wormley 
  - Brad Verhulst
date: Tuesday March 6th 2020
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{Ordinal Data Practical for IBG Boulder Workshop}
  %\usepackage[UTF-8]{inputenc}
---

```{r setup, include = F}
is_CRAN <- !identical(Sys.getenv("NOT_CRAN"), "true")
if (!is_CRAN) {
   options(mc.cores = parallel::detectCores())
} else {
  knitr::opts_chunk$set(eval = FALSE)
  knitr::knit_hooks$set(evaluate.inline = function(x, envir) x)
}
```


```{r start-of-funcs}
intPlot <- function(rho = 0, Tw1Th1 = -Inf, Tw1Th2 = Inf, Tw2Th1 = -Inf, Tw2Th2 = Inf, mini = -3, maxi = 3) {
  den <- function(T1, T2) { # Function to calculate the bivariate normal distribution
    mu1 <- 0 # set the expected value of x1
    mu2 <- 0 # set the expected value of x2
    s11 <- 1 # set the variance of x1
    s22 <- 1 # set the variance of x2
    term1 <- 1 / (2 * pi * sqrt(s11 * s22 * (1 - rho^2)))
    term2 <- -1 / (2 * (1 - rho^2))
    term3 <- (T1 - mu1)^2 / s11
    term4 <- (T2 - mu2)^2 / s22
    term5 <- 2 * rho * ((T1 - mu1) * (T2 - mu2)) / (sqrt(s11) * sqrt(s22))
    term1 * exp(term2 * (term3 + term4 - term5))
  }

  T1 <- round(seq(mini, maxi, by = .1), 2) # generate series for values of x1
  T2 <- round(seq(mini, maxi, by = .1), 2) # generate series for values of x2

  z <- outer(T1, T2, den) # calculate the density values
  dimnames(z) <- list(T1, T2) # give it dim names

  z[T1 < Tw1Th1, ] <- 0 # remove below
  z[, T2 < Tw2Th1] <- 0 # remove below

  z[T1 > Tw1Th2, ] <- 0 # remove above
  z[, T2 > Tw2Th2] <- 0 # remove above

  # This function is the ploting function
  persp(T1, T2, z,
    col = "skyblue1",
    theta = 30, phi = 20, r = 50, d = 0.1, expand = 0.5,
    ltheta = 90, lphi = 180, ticktype = "detailed", nticks = 5, xlab = "Twin 1", ylab = "Twin 2", zlab = "Density"
  )
}



CellProp <- function(thr, r, N) {
  prop <- matrix(omxAllInt(
    matrix(c(1, r, r, 1), 2, 2),
    matrix(0, 1, 2),
    t(matrix(c(rep(-Inf, 2), thr, thr, rep(Inf, 2)), 2, 3))
  ), 2, 2)
  round(N * prop)
}
```



In the first part of the practical we will look at the area under the univariate normal distribution.
The qnorm() function provides the z statistic for a specified probability

```{r part-1}
qnorm(.05)
qnorm(.975)

################
####  Questions  ####
################


# What is the z statistic for a probability of .5?
qnorm(.5)
# > 0

# What is the z statistic for a probability of .99?
qnorm(.99)
# > 2.326348

# What is the z statistic for a probability of .90?
qnorm(.9)
# > 1.281552
```


In the second part of the practical we will used the first function to look at the volume 
under the bivariate normal distribution with various combinations of thresholds and correlations


All of the arguments to the function have defaults, so the function will produce a graph if you just run the line below.

```{r part-2, out.width="80%"}
intPlot()

# Looking at various combinations of affected or unaffected
intPlot(rho = 0, Tw1Th1 = -5, Tw1Th2 = 0, Tw2Th1 = -5, Tw2Th2 = 0) # Concordant unaffected
intPlot(rho = 0, Tw1Th1 = -5, Tw1Th2 = 0, Tw2Th1 = 0, Tw2Th2 = 5) # Discordant: Tw 1 unaffected | Tw 2 affected
intPlot(rho = 0, Tw1Th1 = 0, Tw1Th2 = 5, Tw2Th1 = -5, Tw2Th2 = 0) # Discordant: Tw 1 unaffected | Tw 2 affected
intPlot(rho = 0, Tw1Th1 = 0, Tw1Th2 = 5, Tw2Th1 = 0, Tw2Th2 = 5) # Concordant affected

################
####  Questions  ####
################

# What would the graph look like if there was a .5 correlation between twins?
intPlot(rho = 0)

# What would the graph look like if there was a .5 correlation between twins for concordant unaffected twin pairs?
intPlot(rho = 0, Tw1Th1 = -5, Tw1Th2 = 0, Tw2Th1 = -5, Tw2Th2 = 0) # Concordant unaffected


# What would the graph look like if there was a .5 correlation between twins for concordant affected twin pairs?
intPlot(rho = 0, Tw1Th1 = 0, Tw1Th2 = 5, Tw2Th1 = 0, Tw2Th2 = 5) # Concordant affected



# What would the graph look like if there was a .5 correlation between twins for discordant twin pairs? (Hint this is two graphs)
intPlot(rho = 0, Tw1Th1 = -5, Tw1Th2 = 0, Tw2Th1 = 0, Tw2Th2 = 5) # Discordant: Tw 1 unaffected | Tw 2 affected
intPlot(rho = 0, Tw1Th1 = 0, Tw1Th2 = 5, Tw2Th1 = -5, Tw2Th2 = 0) # Discordant: Tw 1 unaffected | Tw 2 affected
```


# Okay now we want to know how many people we would expect to have in each cell 
```{r part-3}
CellProp(thr = 0, r = 0, N = 1000)


################
####  Questions  ####
################


# Question 6: How many a) Concordant Affected, b) Discordant, c) Concordant Unaffected twin pairs would there be if the correlation was .6?
qnorm(.6)
# > 0.2533471

CellProp(thr = 0, r = 0.2533471, N = 1000)

# a) 291
# b) 291
# c)209

# Question 7: How many a) Concordant Affected, b) Discordant, c) Concordant Unaffected twin pairs would there be if the correlation was .3?

qnorm(.3)
# > -0.5244005

CellProp(thr = 0, r = -0.5244005, N = 1000)
CellProp(thr = -0.5244005, r = 0, N = 1000)

# a) 162
# b) 162
# c) 338
```


# Okay now let's put this into a twin model
```{r part-4}

# First, let's assume that we want to have a model where:
Va <- .33
Vc <- .33
Ve <- .34

# What would we expect the correlations for MZ and DZ twins to be?

rMZ <- Va + Vc
rDZ <- 1/2*Va + Vc

# Second let's assume that 25% of the population has the trait (and 75% does not) (This is in the depression anxiety range).
# What would the thresholds be?


rMZ <- Va + Vc
rDZ <- 1/2*Va + Vc


# Third let's assume that we will have 500 MZ twins and 500 DZ twins.

# How many a) Concordant Affected, b) Discordant, c) Concordant Unaffected MZ twin pairs would we expect?

CellProp(thr = qnorm(1- .25), r = rMZ, N = 500)

nmzConUnaff <- 322
nmzDiscord  <- 53*2
nmzConAff   <- 72

# How many a) Concordant Affected, b) Discordant, c) Concordant Unaffected DZ twin pairs would we expect?
CellProp(thr = qnorm(1- .25), r = rDZ, N = 500)

ndzConUnaff <- 310
ndzDiscord  <- 65*2
ndzConAff   <- 60


# Ok, let's put this into two data frames
mzData <- as.data.frame(rbind(
                matrix(c(0,0), nmzConUnaff, 2, byrow = T),
                matrix(c(1,0), nmzDiscord/2, 2, byrow = T),
                matrix(c(0,1), nmzDiscord/2, 2, byrow = T),
                matrix(c(1,1), nmzConAff, 2, byrow = T)   ))

dzData <- as.data.frame(rbind(
                matrix(c(0,0), ndzConUnaff, 2, byrow = T),
                matrix(c(1,0), ndzDiscord/2, 2, byrow = T),
                matrix(c(0,1), ndzDiscord/2, 2, byrow = T),
                matrix(c(1,1), ndzConAff, 2, byrow = T)   ))

colnames(mzData)  <- colnames(dzData) <- c("obs_T1", "obs_T2")

table(mzData)
table(dzData)


cor(mzData)   # Do these correlations look correct?
rMZ

cor(dzData)
rDZ

# Tell openmx that we are working with ordinal data
mzData <- mxFactor(mzData, c(0:1), labels=c(1,2))
dzData <- mxFactor(dzData, c(0:1), labels=c(1,2))

# Put the data into an mxData object
MZdata  <-  mxData(mzData, type = "raw")
DZdata  <-  mxData(dzData, type = "raw")

# Now let's put this into a twin model and see what comes out
nv <- 1                                                           # Specify the number of variables per twin
ntv <- nv*2                                                       # Specify the total number of variables
selVars <- colnames(mzData)                                       # Specify the variable names

VA <-  mxMatrix( type="Symm", nrow=nv, ncol=nv, free=TRUE, values=c(.1), label=c("VA11"), name="VA" )      # The additive genetic variance component 
VC <-  mxMatrix( type="Symm", nrow=nv, ncol=nv, free=TRUE, values=c(.1), label=c("VC11"), name="VC" )      # The common environmental variance component 
VE <-  mxMatrix( type="Symm", nrow=nv, ncol=nv, free=TRUE, values=c(.8), label=c("VE11"), name="VE" )      # The unique environmental variance component

expMean    <-  mxMatrix( type="Full", nrow=1, ncol=ntv, free=F, values= 0, label=c("mean"), name="expMean" )    # The expected means
Thr        <- mxMatrix("Full", 1, 2, free = T, values = 0, labels = "thr", name = "Thr")                        # The expected thresholds

expCovMZ <- mxAlgebra( expression= rbind  ( cbind(VA+VC+VE , VA+VC), cbind(VA+VC   , VA+VC+VE)), name="expCovMZ" )                     # The MZ expected covariance matrix
expCovDZ <- mxAlgebra( expression= rbind  ( cbind(VA+VC+VE     , 0.5%x%VA+VC), cbind(0.5%x%VA+VC , VA+VC+VE)),  name="expCovDZ" )      # The DZ expected covariance matrix

# Set up the constraint for the variance of the latent liability
vars <- mxAlgebra( VA+VC+VE, name="vars" )                        # calculate the phenotypic variance
con  <- mxConstraint(vars[1,1] ==1, name = "con")              # fix the phenotypic variance to 1


obs <-  list(VA,VC,VE,expMean, Thr, expCovMZ,expCovDZ, vars, con)                                                    # Put all of the objects into a single place
fun <- mxFitFunctionML()                                                                                          # Specify the fit function
mzExp <- mxExpectationNormal(covariance="expCovMZ", means="expMean", thresholds = "Thr", dimnames=selVars )       # Specify the MZ expectation 
dzExp <- mxExpectationNormal(covariance="expCovDZ", means="expMean", thresholds = "Thr", dimnames=selVars )       # Specify the DZ expectation 

MZ <- mxModel("MZ", obs, MZdata, fun, mzExp)                # The MZ model
DZ <- mxModel("DZ", obs, DZdata, fun, dzExp)                # The DZ model

aceFun <- mxFitFunctionMultigroup(c("MZ","DZ"))             # The multiple group fit function
ace <- mxModel("ACE", MZ, DZ, fun, aceFun)                  # The ACE model

aceFit <- mxRun(ace)                                        # Run the model
summary(aceFit)                                             # Look at the summary

# Remember these were the parameters that we initially used
#qnorm(1- .25)

# Va <- .33
# Vc <- .33
# Ve <- .34
```




